% !TeX root = ../main.tex

\chapter{Discussion}\label{chapter:discussion}

\section{Data}
It is clear to see that dietary patterns vary drastically not just between users, but also between days of a user.
This high variance might be due to the user demographic which consists of university students.
Whatever the reason, this lack of easily discernible patterns naturally makes pattern recognition harder.

Another obstacle that makes pattern recognition harder is the apperant lack of data.
There are only 3 users with more than 10 entries, which means that we can train and test models only for these 3 users.

\section{Dialog Agent}
This lack of data for the other 3 users can be attributed to a common feedback about the dialog agent:
Users often forget to log what eat.
This common feedback points to a need for a reminder system, which could be implemented as future work.

Multiple users stated that they usually logged all their meals at the end of the day collectively.
This behaviour obviously leaves most of the predictive functionality obsolete, as the prediction occurs at the time of logging.
One possible solution to this problem is to schedule all the messages for the day at the start of the day, and then optimize the timing as the user makes entries.
This functionality could also work as a reminder, solving two problems at once.

One of the users suggested adding a functionality to display all logged entries, in a similar fashion to \autoref{fig:user-pattern}.
She argued that this could help people see their eating patterns and try to change them for the better.

\section{Machine Learning}
Overall, users stated that the system would somewhat accurately predict when they would get hungry.

Machine learning models achieved suprisingly high $R^2$ scores; over 0.85 for two users and 0.67 for another.
Ensemble learning methods outperformed basic models as expected;
Best performing models were Random Forests for two users, and Bag of Linear Regressors for the other one.
The fact that the best performing models vary between users is an indicator that there isn't a single best model for our task. 

There is another interesting point to mention here:
User 705775462 had the highest entry frequency, followed by user 2075013, followed by user 244717335.
However, while the predictive model for user  705775462 held the highest score, 
it is closely followed by the predictive model for user 244717335.
This seems to indicate that predictibility and eating frequency are not directly related.


